Wrangling multivariate spatio-temporal data with the R package cubble

### 2022-04-30 2-0-cube.Rmd
Updating 2-0-cubble section

Spatio-temporal data can be conceptualised using a cubical data model with three axes, typically time, latitude and longitude. This abstraction can be useful for generalising operations and visualisation purposes: @lu_multidimensional_2018 shows how array operations (select, scale, reduce, rearrange, and compute) can be mapped onto the cube; @bach_review_2014 reviews the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article "does not need to involve spatial data", but refers to "an abstract 2D substrate that is used to visualize data at a specific time". Despite its main focus being on temporal data, the mindset of abstracting out data representation to construct visualisation still applies to our spatio-temporal data manipulation and visualisation.

The most common data cube using the three axes (time, latitude, longitude)  could be considered as taking snapshots of the space and stacking on time. Here we will define the spatio-temporal cube slightly differently with the three axes (time, site, variable), as illustrated in the leftmost cube in Figure \ref{fig:cube-def}. The time axis is the same in both versions, while the site axis now captures both latitude and longitude. Finally, variables are stacked on this space-time canvas, with one observation per site and time point. This notion is adopted to avoid using hypercubes when describing multivariate spatio-temporal data and is the conceptual framework behind \pkg{cubble}. With this conceptual model, operations on spatio-temporal data can be mapped to operations on the cube and the rest of Figure \ref{fig:cube-def} show examples of slicing on site, time, and variable.

While a cubical data model is neat to view the spatio-temporal data conceptually, a 3D data representation, or an array,  may not be convenient for data wrangling. There are two reasons for this: 1) While arrays are efficient for the computation on numerical values, spatio-temporal data contains more than just numerical variables. For example, wrangling on character strings and specific datetime classes is common for spatio-temporal data but may not be easily done with array objects. 2) Creating new variables in an array is more complex than in a data frame since it now needs to be defined in a 2D space, rather than a 1D vector, to be added into the array.

>>>>

Spatio-temporal data can be conceptualised using a cubical data model with three axes which typically are, time, latitude and longitude. This abstraction can be useful for generalising operations and visualisation purposes: @lu_multidimensional_2018 shows how array operations (select, scale, reduce, rearrange, and compute) can be mapped onto the cube; @bach_review_2014 reviews the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article "does not need to involve spatial data", but refers to "an abstract 2D substrate that is used to visualize data at a specific time". Despite its main focus being on temporal data, the mindset of abstracting out data representation to construct visualisations, still applies to our spatio-temporal data manipulation and visualisation approaches.

The most common data cube using the three axes (time, latitude, longitude) could be considered as taking snapshots of the space and stacking them across time. Here we define the spatio-temporal cube slightly differently by considering the three axes to be, time, site and  variable/s. That is illustrated in the leftmost cube in Figure \ref{fig:cube-def}. The time axis is the same in both versions, while the site axis now captures both latitude and longitude. Finally, variables are stacked on this space-time canvas, with one observation per site and time point. This notion is adopted to avoid using hypercubes when describing multivariate spatio-temporal data and is the conceptual framework behind \pkg{cubble}. With this conceptual model, operations on spatio-temporal data can be mapped to operations on the cube and the rest of Figure \ref{fig:cube-def} show examples of slicing on site, time, and variable.

While a cubical data model is neat to view spatio-temporal data conceptually, a 3D data representation, or an array,  may not be convenient for data wrangling. There are two reasons for this: 1) While arrays are efficient for the computation on numerical values, spatio-temporal data, in general, contains more than just numerical variables. For example, wrangling on character strings and specific datetime classes is common for spatio-temporal data but may not be easily done with array objects. 2) Creating new variables in an array is more complex than in a data frame since it would need to be defined in a 2D space, rather than a 1D vector, to be added into the array.

### 2022-04-30 arxiv/cubble-arxiv.tex
Updating 2-0-cubble section

Spatio-temporal data can be conceptualised using a cubical data model with three axes, typically time, latitude and longitude. This abstraction can be useful for generalising operations and visualisation purposes: \protect\hyperlink{ref-lu_multidimensional_2018}{Lu, Appel, and Pebesma} (\protect\hyperlink{ref-lu_multidimensional_2018}{2018}) shows how array operations (select, scale, reduce, rearrange, and compute) can be mapped onto the cube; \protect\hyperlink{ref-bach_review_2014}{Bach et al.} (\protect\hyperlink{ref-bach_review_2014}{2014}) reviews the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article ``does not need to involve spatial data,'' but refers to ``an abstract 2D substrate that is used to visualize data at a specific time.'' Despite its main focus being on temporal data, the mindset of abstracting out data representation to construct visualisation still applies to our spatio-temporal data manipulation and visualisation.

The most common data cube using the three axes (time, latitude, longitude) could be considered as taking snapshots of the space and stacking on time. Here we will define the spatio-temporal cube slightly differently with the three axes (time, site, variable), as illustrated in the leftmost cube in Figure \ref{fig:cube-def}. The time axis is the same in both versions, while the site axis now captures both latitude and longitude. Finally, variables are stacked on this space-time canvas, with one observation per site and time point. This notion is adopted to avoid using hypercubes when describing multivariate spatio-temporal data and is the conceptual framework behind \pkg{cubble}. With this conceptual model, operations on spatio-temporal data can be mapped to operations on the cube and the rest of Figure \ref{fig:cube-def} show examples of slicing on site, time, and variable.

While a cubical data model is neat to view the spatio-temporal data conceptually, a 3D data representation, or an array, may not be convenient for data wrangling. There are two reasons for this: 1) While arrays are efficient for the computation on numerical values, spatio-temporal data contains more than just numerical variables. For example, wrangling on character strings and specific datetime classes is common for spatio-temporal data but may not be easily done with array objects. 2) Creating new variables in an array is more complex than in a data frame since it now needs to be defined in a 2D space, rather than a 1D vector, to be added into the array.

>>>>

Spatio-temporal data can be conceptualised using a cubical data model with three axes which typically are, time, latitude and longitude. This abstraction can be useful for generalising operations and visualisation purposes: \protect\hyperlink{ref-lu_multidimensional_2018}{Lu, Appel, and Pebesma} (\protect\hyperlink{ref-lu_multidimensional_2018}{2018}) shows how array operations (select, scale, reduce, rearrange, and compute) can be mapped onto the cube; \protect\hyperlink{ref-bach_review_2014}{Bach et al.} (\protect\hyperlink{ref-bach_review_2014}{2014}) reviews the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article ``does not need to involve spatial data,'' but refers to ``an abstract 2D substrate that is used to visualize data at a specific time.'' Despite its main focus being on temporal data, the mindset of abstracting out data representation to construct visualisations, still applies to our spatio-temporal data manipulation and visualisation approaches.

The most common data cube using the three axes (time, latitude, longitude) could be considered as taking snapshots of the space and stacking them across time. Here we define the spatio-temporal cube slightly differently by considering the three axes to be, time, site and variable/s. That is illustrated in the leftmost cube in Figure \ref{fig:cube-def}. The time axis is the same in both versions, while the site axis now captures both latitude and longitude. Finally, variables are stacked on this space-time canvas, with one observation per site and time point. This notion is adopted to avoid using hypercubes when describing multivariate spatio-temporal data and is the conceptual framework behind \pkg{cubble}. With this conceptual model, operations on spatio-temporal data can be mapped to operations on the cube and the rest of Figure \ref{fig:cube-def} show examples of slicing on site, time, and variable.

While a cubical data model is neat to view spatio-temporal data conceptually, a 3D data representation, or an array, may not be convenient for data wrangling. There are two reasons for this: 1) While arrays are efficient for the computation on numerical values, spatio-temporal data, in general, contains more than just numerical variables. For example, wrangling on character strings and specific datetime classes is common for spatio-temporal data but may not be easily done with array objects. 2) Creating new variables in an array is more complex than in a data frame since it would need to be defined in a 2D space, rather than a 1D vector, to be added into the array.
